---
title: "Revised Technical File: Conceptualizing Topologies of Antiblackness in Education: A Mathematical Sociology Perspective"
author: 
  - "Nathan N. Alexander, PhD"
  - "Howard University"
#bibliography: brown-references.bib # include references file
#csl: apa.csl
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
  pdf_document:
    toc: false
    toc_depth: 4
    number_section: true 
  word_document:
    toc: false
    reference_docx: word-styles-reference.docx
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo    = knitr::is_html_output(),  # code only in HTML
  results = knitr::is_html_output(),  # results only in HTML  
  warning = FALSE,
  message = FALSE
)

library(here)
here::i_am("technical-revisions-topologies-of-antiblackness.Rmd")
```

# Overview

This study conceptualizes antiblackness as a relational, system-level pattern of separation between Black people and the nonBlack population. It develops comparable Black–nonBlack measures of residential and school segregation across large U.S. metropolitan areas, using a mathematical sociology approach that treats residential and school segregation as linked but distinct distributional systems. This study makes Black–nonBlack separation the analytic object and evaluates whether that separation is reproduced, intensified, or disrupted between neighborhoods and schools.

The study's theoretical object is a conceptualization of antiblackness operationalized as Black–nonBlack separation across residential and school systems. This analysis situates a foundation for future work on hypersegregation across systems.

# Study Design

This study examines Black–nonBlack segregation in residential and educational settings across the largest U.S. Metropolitan Statistical Areas (MSAs). It uses a two-stage analytic design.

First, the study estimates Black–nonBlack residential segregation across Census tracts within each MSA using measures of evenness and exposure. Second, it estimates Black–nonBlack school segregation across public schools within the same MSAs using parallel measures of evenness and exposure. The analysis then compares residential and school segregation profiles to identify metropolitan contexts in which Black separation from the nonBlack population converges, diverges, or compounds across housing and education systems.

The study defines Black residents as Black or African American alone, not Hispanic or Latino. The nonBlack comparison population is all remaining residents or students after removal of the focal Black population. SEDA residential and school segregation files are retained as methodological benchmarks and contextual comparison data; primary Black–nonBlack measures are calculated from tract-level ACS population counts and school-level CCD enrollment counts.

```{r, message=F, warning=F}
# file setup and global options
library(tidyverse)
library(sf)
library(tidycensus)
library(segregation)
library(spdep)
library(knitr)
library(ggplot2)
library(forcats)
library(purrr)
library(janitor)
options(tigris_class = "sf", 
        tigris_use_cache = TRUE)
```

# Load data

In this section, I load the study data. Residential segregation benchmark measures are drawn from SEDA’s 2018–2022 ACS-based estimates, and school segregation benchmark measures are drawn from SEDA’s 2021–2022 school-year data. Contextual tract characteristics are drawn from the 2018–2022 ACS five-year estimates.

Residential, school, and tract-context data are selected from the closest available periods so that observed residential–school differences are interpreted as contemporary cross-system patterns rather than artifacts of comparing substantially different historical periods.

SEDA data are retained as benchmark and validation measures. The study’s primary Black–nonBlack residential and school segregation measures will be calculated from tract- and school-level population or enrollment data using the measurement conventions established in the SEDA documentation where applicable.

## SEDA data

### Residential

`residential_seg_cbsa23_1.0`

This file provides metropolitan-level residential segregation measures using the 2023 CBSA geographic definition. This SEDA workbook is downloaded to a temporary session file, imported into memory, and deleted immediately after import.

```{r, warning=F, message=F}
res_seg_url <- paste0(
  "https://stacks.stanford.edu/file/druid:",
  "mg998jk9421/residential_seg_cbsa23_1.0.csv"
)

options(timeout = 600)  # 10 minutes

tmp <- tempfile(fileext = ".xlsx")

download.file(
  res_seg_url,
  destfile = tmp,
  method = "libcurl",
  mode = "wb",
  quiet = FALSE
)

res_seg <- readxl::read_excel(tmp, sheet = 1)

unlink(tmp)
```

#### Check: Residential

```{r}
names(res_seg)

dplyr::count(
  res_seg, 
  year, 
  source)

dplyr::n_distinct(res_seg$cbsa23id)

dplyr::count(
  res_seg, 
  year, 
  cbsa23id
  ) %>%
  dplyr::filter(n > 1)

res_seg %>%
  count(year, cbsa23id) %>%
  summarise(
    duplicated_cbsa_years = sum(n > 1),
    maximum_rows_per_cbsa_year = max(n)
  )

res_seg %>%
  count(year) %>%
  arrange(year)

dplyr::glimpse(res_seg)

```

### School

`school_seg_cbsa23_2.0`

This file provides metropolitan-level school segregation measures estimated across schools and related school-system units.

```{r}

sch_seg_url <- paste0(
  "https://stacks.stanford.edu/file/druid:gh906mk5473/school_seg_cbsa23_2.0.csv"
)

options(timeout = 600)  # 10 minutes

tmp <- tempfile(fileext = ".xlsx")

download.file(
  sch_seg_url,
  destfile = tmp,
  method = "libcurl",
  mode = "wb",
  quiet = FALSE
)

sch_seg <- readxl::read_excel(tmp, sheet = 1)

unlink(tmp)
```

#### Check: School
```{r}
names(sch_seg)

dplyr::glimpse(sch_seg)

names(sch_seg)[
  stringr::str_detect(
    names(sch_seg),
    stringr::regex(
      "year|source|cbsa|tot|blk|wht|hsp|asn|per|n",
      ignore_case = TRUE
    )
  )
]

```

### LISD 

The Longitudinal Imputed School Dataset (LISD) and SEDA Geographic Crosswalk provide the cleaned, imputed, and geographically harmonized school data used in SEDA’s school-segregation estimates. 

LISD will be retained alongside the raw CCD data retrieved through the Urban Institute to allow comparison of school-level counts and selection of the most appropriate data source for the study’s Black–nonBlack measures.

```{r, warning=F, message=F}
lisd_url <- paste0(
  "https://stacks.stanford.edu/file/druid:",
  "gh906mk5473/LISD_2.0_2018-2023.csv"
)

options(timeout = 600)  # 10 minutes

tmp <- tempfile(fileext = ".xlsx")

download.file(
  lisd_url,
  destfile = tmp,
  method = "libcurl",
  mode = "wb",
  quiet = FALSE
)

lisd <- readxl::read_excel(tmp, sheet = 1)

unlink(tmp)
```

#### Check: LISD

```{r}
names(lisd)

lisd %>%
  dplyr::count(year) %>%
  dplyr::arrange(year)

lisd %>%
  dplyr::summarise(
    n_school_year_records = dplyr::n(),
    n_schools = dplyr::n_distinct(nces7_final),
    n_original_ncessch = dplyr::n_distinct(ncessch),
    n_missing_member = sum(is.na(member)),
    n_missing_black = sum(is.na(tot_black)),
    n_missing_type = sum(is.na(type)),
    n_missing_virtual = sum(is.na(virtual))
  )

lisd %>%
  dplyr::count(
    year,
    nces7_final,
    name = "n_records"
  ) %>%
  dplyr::filter(n_records > 1)
```

For LISD:

    member is the total school membership/enrollment measure.

    tot_black is Black enrollment.

    tot_nam, tot_asian, tot_hisp, tot_tr, and tot_white are the other reported race/ethnicity counts.

    grdlo and grdhi identify the lowest and highest grades served, but LISD does not provide enrollment separately by grade.

    type provides the school-type label; your displayed entries are "Regular School".

    virtual is numeric, so inspect its codes before using it as an eligibility filter.

    cbsa23id and cbsa23name are the crosswalk-derived 2023 CBSA fields.

    metrostatus23 is already "Metropolitan area" because you created lisd_metro using that restriction.
    
### Geographic crosswalk

The Geographic Crosswalk will be used to identify each school’s 2023 CBSA/MSA membership. The SEDA school-segregation workbook remains a CBSA-level benchmark for evaluating comparable measures.


```{r, warning=F, message=F}
geo_crosswalk_url <- paste0(
  "https://stacks.stanford.edu/file/druid:",
  "gh906mk5473/geo_crosswalk_2.0_2018-2023.csv"
)

options(timeout = 600)  # 10 minutes

tmp <- tempfile(fileext = ".xlsx")

download.file(
  geo_crosswalk_url,
  destfile = tmp,
  method = "libcurl",
  mode = "wb",
  quiet = FALSE
)

geo_crosswalk <- readxl::read_excel(tmp, sheet = 1)

unlink(tmp)
```

#### Check: Geocrosswalk

```{r}
names(geo_crosswalk)
```
### Join LISD-GEO

```{r}
lisd_geo <- lisd %>%
  dplyr::left_join(
    geo_crosswalk %>%
      dplyr::select(
        year,
        nces7_final,
        CBSACode23,
        CBSATitle23,
        metrostatus23
      ),
    by = c(
      "year",
      "nces7_final"
    )
  )

# rename to match to SEDA CBSA benchmark
lisd_geo <- lisd_geo %>%
  dplyr::rename(
    cbsa23id = CBSACode23,
    cbsa23name = CBSATitle23,
    metrostatus23 = metrostatus23
  )
```

Check: JOIN

```{r}
# confirm one record for each school-year

geo_crosswalk %>%
  dplyr::count(
    year,
    nces7_final,
    name = "n_crosswalk_records"
  ) %>%
  dplyr::filter(n_crosswalk_records > 1)

# calculate join coverage

lisd_geo %>%
  dplyr::summarise(
    n_lisd_records = dplyr::n(),
    n_matched_crosswalk = sum(!is.na(cbsa23id)),
    pct_matched_crosswalk = mean(!is.na(cbsa23id)) * 100,
    n_unmatched_crosswalk = sum(is.na(cbsa23id))
  )

# inspect unmatched records
lisd_geo %>%
  dplyr::filter(is.na(cbsa23id)) %>%
  dplyr::select(
    year,
    nces7_final,
    ncessch,
    ncessch_orig,
    schoolname,
    fips
  ) %>%
  dplyr::arrange(year, fips, schoolname)

# diagnose unmatched CBSA code
lisd_geo_check <- lisd %>%
  dplyr::left_join(
    geo_crosswalk %>%
      dplyr::transmute(
        year,
        nces7_final,
        crosswalk_record_found = TRUE,
        cbsa23id = CBSACode23,
        cbsa23name = CBSATitle23,
        metrostatus23
      ),
    by = c(
      "year",
      "nces7_final"
    )
  )

lisd_geo_check %>%
  dplyr::count(
    crosswalk_record_found,
    cbsa23id_missing = is.na(cbsa23id),
    metrostatus23,
    sort = TRUE
  )

# Create school-year LISD

lisd_metro <- lisd_geo_check %>%
  dplyr::filter(
    metrostatus23 == "Metropolitan area"
  )
```

## ACS data

5-year estimates (2018–2022)

```{r, message=F, warning=F}
acs_year <- 2022  # 2018–2022 ACS 5-year estimates

cat(
  "Using",
  acs_year,
  "ACS 5-year estimates (2018–2022).\n"
)

options(
  tigris_class = "sf",
  tigris_use_cache = TRUE
)
```

### Race variables

```{r, message=F, warning=F}
# National tract-level ACS data for the largest U.S. MSAs
# ACS 2018–2022 5-year estimates
# Hispanic or Latino Origin by Race: B03002
#
# Race/ethnicity variables are drawn from ACS Table B03002:
#   Pop      = B03002_001 (Total population)
#   White    = B03002_003 (Not Hispanic White alone)
#   Black    = B03002_004 (Not Hispanic Black alone)
#   Asian    = B03002_006 (Not Hispanic Asian alone)
#   Hispanic = B03002_012 (Hispanic or Latino, any race)
#
# Study comparison:
#   Black    = Black alone, not Hispanic or Latino
#   NonBlack = Total population - Black

acs_vars <- c(
  totalpop = "B03002_001",
  white    = "B03002_003",
  black    = "B03002_004",
  asian    = "B03002_006",
  hispanic = "B03002_012"
)
```

### MSA reference geography

#### Load 2023 CBSA/MSA reference geography

```{r, message=F, warning=F}
cbsa_2023 <- tigris::core_based_statistical_areas(
  year = 2023,
  class = "sf"
)

names(cbsa_2023)

dplyr::glimpse(cbsa_2023)
```

#### Create MSA lookup

The analytic unit is the Metropolitan Statistical Area. The SEDA workbooks use 2023 CBSA identifiers; MSAs are the metropolitan subset of CBSAs. This section creates the 2023 MSA reference geography and county-to-MSA lookup used to assign tract-level ACS data to the MSA analytic units. Micropolitan CBSAs are excluded.

```{r}
msa_lookup <- cbsa_2023 %>%
  dplyr::filter(
    LSAD == "M1"
  ) %>%
  dplyr::transmute(
    cbsa23id = CBSAFP,
    cbsa23name = NAME,
    csa = CSAFP,
    cbsa_land_area_sq_m = ALAND,
    cbsa_water_area_sq_m = AWATER,
    geometry
  )

msa_lookup %>%
  dplyr::summarise(
    n_msas = dplyr::n(),
    duplicated_cbsa23id = sum(duplicated(cbsa23id))
  )
```

#### Download List 1 county-to-CBSA file

```{r}
county_cbsa_url <- paste0(
  "https://www2.census.gov/programs-surveys/metro-micro/",
  "geographies/reference-files/2023/delineation-files/",
  "list1_2023.xlsx"
)

tmp <- tempfile(fileext = ".xlsx")

download.file(
  county_cbsa_url,
  destfile = tmp,
  method = "libcurl",
  mode = "wb",
  quiet = FALSE
)

readxl::excel_sheets(tmp)

county_cbsa_raw <- readxl::read_excel(
  tmp,
  sheet = 1,
  skip = 2
)

unlink(tmp)

names(county_cbsa_raw)

dplyr::glimpse(county_cbsa_raw)
```

#### Create county_msa_lookup

```{r}
county_msa_lookup <- county_cbsa_raw %>%
  dplyr::filter(
    `Metropolitan/Micropolitan Statistical Area` ==
      "Metropolitan Statistical Area"
  ) %>%
  dplyr::transmute(
    county_geoid = paste0(
      `FIPS State Code`,
      `FIPS County Code`
    ),
    cbsa23id = `CBSA Code`,
    cbsa23name = `CBSA Title`,
    central_outlying_county = `Central/Outlying County`
  )
```

### Tract-level ACS data

In this section, I retrieve ACS 2018–2022 tract-level population estimates and tract geometry for the national MSA universe. The tract file will supply the Black and nonBlack population counts, geometry, and land-area inputs needed to calculate residential segregation dimensions and to create maps of selected MSAs.

#### Download ACS tract counts and geometry

```{r}
state_fips <- c(
  state.abb,
  "DC"
)

tract_data_raw <- purrr::map_dfr(
  state_fips,
  function(state_code) {
    tidycensus::get_acs(
      geography = "tract",
      variables = acs_vars,
      year = acs_year,
      survey = "acs5",
      state = state_code,
      geometry = TRUE,
      output = "wide"
    )
  }
)
```

#### Create tract_data with Black/nonBlack variables and density fields

```{r}
# Zero-area tract safeguard
tract_data <- tract_data_raw %>%
  dplyr::transmute(
    GEOID,
    NAME,
    county_geoid = stringr::str_sub(
      GEOID,
      start = 1,
      end = 5
    ),
    totalpop = totalpopE,
    white = whiteE,
    black = blackE,
    asian = asianE,
    hispanic = hispanicE,
    nonblack = totalpop - black,
    geometry
  ) %>%
  sf::st_transform(5070) %>%
  dplyr::mutate(
    land_area_sq_m = as.numeric(
      sf::st_area(geometry)
    ),
    land_area_sq_km = land_area_sq_m / 1000000,
    black_population_density_sq_km = dplyr::if_else(
      land_area_sq_km > 0,
      black / land_area_sq_km,
      NA_real_
    )
  )
```

### Assign tracts to MSAs

Census tracts are linked to MSAs using the county-to-CBSA relationship file associated with the 2023 OMB delineations. This follows SEDA’s approach of linking tracts to CBSAs through counties.

#### Create tract_data_msa

```{r}
tract_data_msa <- tract_data %>%
  dplyr::left_join(
    county_msa_lookup %>%
      dplyr::select(
        county_geoid,
        cbsa23id,
        cbsa23name
      ),
    by = "county_geoid"
  ) %>%
  dplyr::semi_join(
    msa_lookup %>%
      sf::st_drop_geometry() %>%
      dplyr::select(cbsa23id),
    by = "cbsa23id"
  )
```

### Select largest MSAs

The analytic sample consists of the 100 largest U.S. MSAs, ranked by total population in the 2018–2022 ACS five-year estimates.

#### Set n_largest_msas

```{r}
n_largest_msas <- 100
```

#### Create msa_population

```{r}
msa_population <- tract_data_msa %>%
  sf::st_drop_geometry() %>%
  dplyr::group_by(
    cbsa23id,
    cbsa23name
  ) %>%
  dplyr::summarise(
    msa_totalpop = sum(
      totalpop,
      na.rm = TRUE
    ),
    msa_black = sum(
      black,
      na.rm = TRUE
    ),
    msa_nonblack = sum(
      nonblack,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  dplyr::arrange(
    dplyr::desc(msa_totalpop)
  ) %>%
  dplyr::mutate(
    msa_population_rank = dplyr::row_number()
  )
```

#### Create largest_msas

```{r, message=F, warning=F}
largest_msas <- msa_population %>%
  dplyr::slice_head(
    n = n_largest_msas
  )
```

#### Create tract_data_largest_msas

```{r}
tract_data_largest_msas <- tract_data_msa %>%
  dplyr::semi_join(
    largest_msas %>%
      dplyr::select(cbsa23id),
    by = "cbsa23id"
  )
```

## CCD school data

The 2021–2022 NCES Common Core of Data Public School Universe Survey provides school-level enrollment and location data. These data are used to calculate Black–nonBlack school evenness and exposure within the selected MSAs.

The SEDA school workbook provides CBSA-level benchmark measures and aggregate enrollment counts. The CCD school data provide the school-level enrollment distribution required to calculate the study’s primary Black–nonBlack school segregation measures.

Package notes are available [here](https://urbaninstitute.github.io/education-data-package-r/).

```{r, eval=F, message=F, warning=F}
# Urban Institute's education data package
# install.packages("educationdata")

library(educationdata)

ccd_year <- 2021  # 2021–2022 school year

ccd_directory <- educationdata::get_education_data(
  level = "schools",
  source = "ccd",
  topic = "directory",
  filters = list(
    year = ccd_year
  ),
  add_labels = TRUE
)

ccd_grades <- c(
  "grade-k",
  "grade-1",
  "grade-2",
  "grade-3",
  "grade-4",
  "grade-5",
  "grade-6",
  "grade-7",
  "grade-8",
  "grade-9",
  "grade-10",
  "grade-11",
  "grade-12"
)

ccd_enrollment_race <- educationdata::get_education_data(
  level = "schools",
  source = "ccd",
  topic = "enrollment",
  subtopic = list("race"),
  filters = list(
    year = ccd_year,
    grade = ccd_grades
  ),
  add_labels = TRUE,
  csv = TRUE
)

ccd_enrollment_wide <- ccd_enrollment_race %>%
  dplyr::filter(
    sex == "Total"
  ) %>%
  dplyr::select(
    year,
    ncessch,
    grade,
    race,
    enrollment
  ) %>%
  tidyr::pivot_wider(
    names_from = race,
    values_from = enrollment,
    names_prefix = "enrollment_"
  )

ccd_school_data <- ccd_directory %>%
  dplyr::transmute(
    year,
    ncessch,
    school_name,
    leaid,
    lea_name,
    state_location,
    city_location,
    county_code,
    cbsa23id = stringr::str_pad(
      as.character(cbsa),
      width = 5,
      side = "left",
      pad = "0"
    ),
    latitude,
    longitude,
    school_level,
    school_type,
    school_status,
    lowest_grade_offered,
    highest_grade_offered,
    charter,
    magnet,
    shared_time,
    virtual,
    total_enrollment_directory = enrollment,
    free_lunch,
    reduced_price_lunch,
    free_or_reduced_price_lunch
  ) %>%
  dplyr::left_join(
    ccd_enrollment_wide,
    by = c(
      "year",
      "ncessch"
    )
  ) %>%
  dplyr::mutate(
    total_enrollment = enrollment_Total,
    black_enrollment = enrollment_Black,
    nonblack_enrollment = total_enrollment - black_enrollment
  )
```

#### Data check

```{r, eval=F}
ccd_school_data %>%
  dplyr::summarise(
    n_rows = dplyr::n(),
    n_school_grade_records = dplyr::n_distinct(
      ncessch,
      grade
    ),
    duplicated_school_grade_records = sum(
      duplicated(
        dplyr::pick(
          ncessch,
          grade
        )
      )
    ),
    missing_cbsa23id = sum(is.na(cbsa23id)),
    missing_total_enrollment = sum(is.na(total_enrollment)),
    missing_black_enrollment = sum(is.na(black_enrollment)),
    negative_nonblack_enrollment = sum(
      nonblack_enrollment < 0,
      na.rm = TRUE
    ),
    zero_total_enrollment = sum(
      total_enrollment == 0,
      na.rm = TRUE
    )
  )
```

#### Failure check

```{r, eval=F}
ccd_school_data %>%
  dplyr::filter(
    nonblack_enrollment < 0
  ) %>%
  dplyr::select(
    ncessch,
    school_name,
    state_location,
    city_location,
    total_enrollment,
    black_enrollment,
    nonblack_enrollment
  )
```

# Construct analysis inputs

## Residential input

```{r}
residential_input <- tract_data_largest_msas %>%
  dplyr::filter(
    !is.na(totalpop),
    !is.na(black),
    !is.na(nonblack),
    totalpop > 0,
    black >= 0,
    nonblack >= 0
  ) %>%
  dplyr::select(
    cbsa23id,
    cbsa23name,
    GEOID,
    totalpop,
    black,
    nonblack,
    geometry
  )
```

### School input

The school analytic universe is designed to approximate SEDA’s focus on open, regular, non-virtual schools using the available CCD directory classifications. Because SEDA’s estimates are calculated from LISD, which includes cleaned and imputed CCD data, the primary calculations are treated as a Black–nonBlack adaptation informed by, rather than an exact replication of, SEDA’s school-segregation estimates.

The school analytic file retains open, regular schools with valid grade-specific enrollment counts. Schools explicitly designated as virtual or as virtual with face-to-face options are excluded. Schools coded as offering supplemental virtual instruction and schools with missing virtual-status information are retained because those categories do not identify schools as exclusively virtual; the `virtual` designation is retained for sensitivity analysis.

```{r}
school_input <- ccd_school_data %>%
  dplyr::filter(
    school_status == "Open",
    school_type == "Regular school",
    !virtual %in% c(
      "Yes",
      "Virtual with face-to-face options"
    ),
    !is.na(grade),
    !is.na(total_enrollment),
    !is.na(black_enrollment),
    !is.na(nonblack_enrollment),
    total_enrollment > 50,
    black_enrollment >= 0,
    nonblack_enrollment >= 0
  ) %>%
  dplyr::semi_join(
    largest_msas %>%
      dplyr::select(cbsa23id),
    by = "cbsa23id"
  ) %>%
  dplyr::select(
    cbsa23id,
    year,
    grade,
    ncessch,
    school_name,
    leaid,
    virtual,
    total_enrollment,
    black_enrollment,
    nonblack_enrollment,
    free_or_reduced_price_lunch
  )
```

#### Check: school input

```{r}
school_input %>%
  dplyr::group_by(
    cbsa23id,
    grade
  ) %>%
  dplyr::summarise(
    n_schools = dplyr::n_distinct(ncessch),
    total_enrollment = sum(total_enrollment),
    black_enrollment = sum(black_enrollment),
    nonblack_enrollment = sum(nonblack_enrollment),
    .groups = "drop"
  ) %>%
  dplyr::filter(
    black_enrollment == 0 |
      nonblack_enrollment == 0
  )
```


